\section{The Zermelo-Fraenkel Axioms and the Axiom of Choice}

We begin by reviewing the Zermelo-Fraenkel Axioms of Set Theory. We work in a first-order language with only one relation symbol, denoted $\in$. In principle, we would want to distinguish between the symbol in the language and its interpretation in any model, but we will not do this in practice.

We begin with the Axiom of Extensionality.

\begin{baxiom}[The Axiom of Extensionality]\label{ZFC:Ext}
    \begin{align*}
        \forall A \, \forall B \, \brac{\parenth{\forall x \parenth{x \in A \lr x \in B}} \to \parenth{A = B}}
    \end{align*}
\end{baxiom}

We define the $\subseteq$ symbol, used infix as $A \subseteq B$, to be shorthand for the formula $\forall x \, \parenth{x \in A \to x \in B}$. \Cref{ZFC:Ext} tells us that
\begin{align*}
    \forall A \forall B \parenth{\parenth{A \subseteq B \land B \subseteq A} \to A = B}
\end{align*}

We now define the Axiom Scheme of Comprehension.

\begin{baxiom}[The Axiom (Scheme) of Comprehension]\label{ZFC:Compr}
    Let $x$ and $y$ be free variables. For all formulae $\varphi\of{x, y}$, the following is an axiom:
    \begin{align*}
        \forall A \forall y \exists B \forall x \brac{x \in B \lr \parenth{x \in A \land \varphi\of{x, y}}}
    \end{align*}
    Intuitively, this means we can define
    \begin{align*}
        B = \setst{x \in A}{\varphi\of{x, y}}
    \end{align*}
    This is an axiom scheme because we can increase the arity of $\varphi$ and have more free variables $y_1, y_2, y_3, \ldots, y_n$ for any $n$.
\end{baxiom}

Note that \Cref{ZFC:Ext} tells us that we can replace the existential quantifier for $B$ in \Cref{ZFC:Compr} with an existence and uniqueness quantifier without changing its meaning.

Before proceeding further, we define the empty set. First, observe that in first order logic, structures are required to have non-empty universes. Thus, for every structure $\parenth{A, E}$ in the language of set theory, with $E \subseteq \parenth{A \times A}$ representing equality, we must have
\begin{align*}
    \parenth{A, E} \models \exists x \parenth{x = x}
\end{align*}
The Completeness Theorem then tells us that
\begin{align*}
    \vdash \exists x \parenth{x = x}
\end{align*}
ie, that the formula $\exists x \parenth{x = x}$ \textit{must be a theorem} in the language of sets. Hence, the axioms \Cref{ZFC:Ext,ZFC:Compr} tell us that
\begin{align*}
    \vdash \exists! A \, \forall x \, \parenth{x \notin A}
\end{align*}
In other words, it is a theorem in the language of sets that there is a unique set that contains no members whatsoever. This unique set is denoted $\emptyset$, and is called the \textbf{empty set}. Its existence (and uniqueness) is not a distinct axiom, but a direct consequences of \Cref{ZFC:Ext,ZFC:Compr}.

Next, we give the axiom of pairing.

\begin{baxiom}[The Axiom of Pairing]\label{ZFC:Pairing}
    \begin{align*}
        \forall x \, \forall y \, \exists A \, \parenth{x \in A \land y \in A}
    \end{align*}
\end{baxiom}

\Cref{ZFC:Pairing} essentially gives us a way of constructing, for any $x$ and $y$, the unique set $A = \set{x, y}$ with the property that
\begin{align*}
    \forall z \parenth{z \in A \lr \parenth{z = x} \lor \parenth{z = y}}
\end{align*}
In particular, it allows us to construct, for any $x$, the set $\set{x, x}$, which, by \Cref{ZFC:Ext}, is exactly $\set{x}$. In other words, it tells us that we can stick any $x$ into a set that only contains $x$.

In a similar flavour, we can construct a \textit{union} of any family of sets. Note that this is distinct from pairing, because to construct $\set{x, y}$ as a union of $\set{x}$ and $\set{y}$, one needs to know that one can construct $\set{x}$ and $\set{y}$ in the first place. This requires pairing (or some version of it).

Now, we starte the axiom of unions.

\begin{baxiom}[The Axiom of Unions]
    If we denote by $\calF$ a family of sets,
    \begin{align*}
        \forall \calF \, \exists B \, \forall A \, \brac{A \in \calF \to A \subseteq B}
    \end{align*}
\end{baxiom}

As usual, we can use \Cref{ZFC:Ext,ZFC:Compr} to show that
\begin{align*}
    \bigcup \calF = \setst{x}{\exists A \subseteq \calF \, \parenth{x \in A}}
\end{align*}
is well (and uniquely) defined.

Next comes the power set axiom.

\begin{baxiom}[The Axiom of Power Sets]
    \begin{align*}
        \forall A \, \exists \calF \, \forall X \, \brac{X \subseteq A \to X \in \calF}
    \end{align*}
\end{baxiom}

Again, \Cref{ZFC:Ext,ZFC:Compr} tell us that this is equivalent to
\begin{align*}
    \forall A \, \exists! \calF \, \forall X \, \parenth{X \in \calF \lr X \subseteq A}
\end{align*}
This justifies the definition
\begin{align*}
    \Powset(A) := \setst{X}{X \subseteq A}
\end{align*}

Next comes the second (and last) axiom scheme in ZFC, the axiom scheme of replacement.

\begin{baxiom}[The Axiom (Scheme) of Replacement]\label{ZFC:Replacement}
    For each formula $\varphi\of{x, y, z}$ that does not contain $B$ as a free variable, the following is an axiom:
    \begin{align*}
        \forall A \, \forall z \, \brac{
            \brac{\forall x \in A \, \exists! y \, \varphi\of{x, y, z}}
            \to
            \brac{\brac{\exists B \, \forall x \in A \, \exists y \in B \, \varphi\of{x, y, z}}}
        }
    \end{align*}
    As in \Cref{ZFC:Compr}, we can introduce more free variables in lieu of $z$.
\end{baxiom}

We can unpack this slightly complicated formula using \Cref{ZFC:Ext,ZFC:Compr}: it essentially tells us that we can define
\begin{align*}
    \setst{y}{\exists x \in A \, \varphi\of{x, y, z}}
\end{align*}
The quantifiers $\forall A$ and $\forall z$ merely introduce variables. What the rest of the axiom (scheme) tells us is that \textbf{if} for every $x$, there is a unique \textit{witness} $y$ for which the formula $\varphi\of{x, y, z}$ holds, \textbf{then} we can collect all these witnesses into a set that is contained $B$, ranging over $x \in A$. Intuitively, ``the number of such witnesses is \textit{small enough} that it \textit{can be collected into a set}.''

Indeed, if it looks like $\varphi\of{x, y, z}$ defines a function with domain $A$, then it really does! If we can define $A \times B = \setst{\parenth{x, y}}{x \in A \text{ and } y \in B}$ (which we have yet to show we can do, but let's say we can), then we can put
\begin{align*}
    f = \setst{\parenth{x, y} \in A \times B}{\varphi\of{x, y, z}}
\end{align*}
which is a set by \Cref{ZFC:Compr}. Effectively, \Cref{ZFC:Replacement} tells us that \textit{if the domain of a function is a set, then so is its range}: everything in its ``image'' (a \textit{witness}) can be collected into some set. Of course, we only include this comment for the sake of intuition; as far as the formalism goes, we cannot talk about domains and ranges of functions yet, because functions are not yet defined. That being said, \Cref{ZFC:Replacement} is an essential step in the process of actually \textit{defining} functions (or showing that this is something we can \textit{do}).

One situation in which \Cref{ZFC:Replacement} is used (and will be used once sufficiently many axioms are stated) is in the situation we talked about at the start of \Cref{Ch1:CH}. We'll see that there is a formula $\varphi\of{n, S}$ so that $\varphi\of{n, S} \lr \parenth{n \in \omega \land S = V_n}$. Then, \Cref{ZFC:Replacement} would imply that
\begin{align*}
    \grpres{V_n}{n < \omega}
\end{align*}
really is a sequence. This allows us to define $V_\omega = \bigcup_{n < \omega} V_n$ and continue.

% Next time, we will look at the Foundation Axiom.

% 29 August

We now state the most interesting of the ZFC axioms: the Axiom of Choice.

\begin{baxiom}[The Axiom of Choice - (AC)]\label{ZFC:AC}
    \hfill
    \begin{align*}
        \forall \calF \, \exists c \, \brac{\forall A \in \calF \, \parenth{A \neq \emptyset \to A \in \dom{c} \land c(A) \in A}}
    \end{align*}
    where $\calF$ is a family of sets, $c$ is a function, and $\dom{c}$ is the domain of $c$. We call $c$ a \textbf{Choice Function on $\calF$}.
\end{baxiom}

The reason we state the Axiom of Choice at this point in our discussion is that modulo the ZFC Axioms that we have seen so far, the Axiom of Choice is equivalent to each of the following.
\begin{enumerate}
    \item $\forall A \, \exists R \subseteq A \times A \, \parenth{A, R}$ is a well-ordering.
    \item $\forall A \, \exists \alpha \, \exists f : A \xrightarrow{\sim} \alpha$, where $\alpha$ is an ordinal.
\end{enumerate}

Note that none of the above axioms imply the existence of the natural numbers, without which virtually none of the mathematics we know and love would be possible. We therefore come to the axiom of infinity, which posits the existence of infinite sets, the `smallest' of which is the natural numbers.

\begin{baxiom}[The Axiom of Infinity]\label{ZFC:Infty}
    \hfill
    \begin{align*}
        \exists I \, \parenth{\emptyset \in I \land \forall x \in I \, \parenth{x \cup \set{x} \in I}}
    \end{align*}
    A set $I$ as above is called an \textbf{inductive set}, and it is an axiom of ZFC that such a set exists.
\end{baxiom}

It is possible to prove, with relative ease, that given the existence of inductive sets (as posited above), there is a unique inductive set $\omega$ such that for all inductive sets $J$, $\omega \subseteq J$. Indeed, for any inductive set $K$, $\omega$ can be defined as the intersection of all inductive subsets $J \subseteq K$. It can then be shown that this intersection $\omega$ lives in all inductive sets. This minimal inductive set $\omega$ is known as the \textbf{set of natural numbers}. It is easy to see that the famed axioms of Peano are satisfied by $\omega$.

Recall the sets $V_0, V_1, \ldots, V_\omega, \ldots$ defined at the start of \Cref{Ch1:CH}. It is worth asking ourselves the following question: \textit{Why does $V_{\omega}$ exist?} Or, more precisely, \textit{why is $V_{\omega}$ a set?}

By induction, we know that each $V_n$ is a set. We want to essentially see that the map $n \mapsto V_n$ is a \textit{set} (ie, a \textit{function}). Then, if we take the union of the range of this function, whose domain is the set $\omega$ whose existence is given by \Cref{ZFC:Infty}, we can define $V_{\omega}$ to be that union.

We invoke \Cref{ZFC:Replacement}, the Axiom of Replacement. Let $\varphi\of{x, y}$ be the formula capturing the following facts:
\begin{enumerate}
    \item $y$ is a function with domain $\dom{y} = x \cup \set{x}$.
    \item If $\emptyset \in \dom{y}$, then $y\of{\emptyset} = \emptyset$.
    \item If $z \cup \set{z} \in \dom{y}$, then $z \in \dom{y}$ and $y\of{z \cup \set{z}} = \Powset\of{z}$.
\end{enumerate}
The idea is that $y = \cycl{V_0, V_1, V_2, \ldots, V_{n+1}}$\footnote{By this, we mean it is a \textit{sequence}, ie, a function from $\omega$ to something that takes $0$ to $V_{0}$, $1$ to $V_{1}$, and so on.} for some $n \in \omega$. Then,
\begin{align*}
    \ZFC \vdash \forall n \in \omega \ \exists! y \, \varphi\of{n, y}
\end{align*}
Applying the Axiom of Replacement (\Cref{ZFC:Replacement}), we know that the map that takes any $n \in \omega$ to its unique witness $y$ for $n$ \textit{is, indeed, a function}. In other words,
\begin{align*}
    \grpres{\grpres{V_m}{m \leq n}}{n < \omega}
\end{align*}
is a set. From this, we can read off
\begin{align*}
    \grpres{V_n}{n < \omega}
\end{align*}
by taking a union. % OF WHAT?

\begin{remark}
    Recall that $\parenth{V_{\omega}, \in}$ is shorthand for $\parenth{V_{\omega}, R}$, where $R = \setst{\parenth{x, y} \in V_{\omega} \times V_{\omega}}{x \in y}$. It is possible to show not only that
    \begin{align*}
        \parenth{V_{\omega}, \in} &\models \text{ZFC without the Axiom of Infinity}
    \end{align*}
    but the stronger fact that
    \begin{align*}
        \parenth{V_{\omega}, \in} &\models \text{ZFC without the Axiom of Infinity but with its \textit{negation}}
    \end{align*}
    Thus,
    \begin{align*}
        \ZFC \vdash \text{ZFC without Infinity but with $\neg\text{Infinity}$ is consistent}
    \end{align*}
    Ie, from ZFC, we can deduce that the entirety of ZFC, except with the negation of the axiom of infinity instead of the axiom itself, is consistent.
\end{remark}

We note that the typical background setting for this course is the entirety of ZFC. However, as and when we need to either get rid of an axiom, or introduce a new one, we will mention it explicitly. We also use the following notational conventions.

\begin{boxnotation}
    We use $\ZFC$ to mean we have all the ZFC axioms. As and when we need variations, we do the following.
    \begin{align*}
        \ZF &:= \text{``ZFC without the Axiom of Choice''} \\
        \ZFCwoF &:= \text{``ZFC without the Axiom of Foundation''} \\
        \ZFwoF &:= \text{``ZF without the Axiom of Foundation''}
    \end{align*}
\end{boxnotation}

Here, we end our discussion on the axioms of set theory, and proceed to consider a broader type of object than sets, known as classes.