%% WEEK 1

\chapter{A Recap of Undergraduate Set Theory} \label{Ch1:CH}
\thispagestyle{empty}

Set Theory can be viewed as a field of mathematics that encompasses all other mathematics. We can view set theory as beginning at the empty set $\emptyset$, also denoted $V_0$. We can construct subsequent sets $V_1, V_2, V_3, \ldots$ by taking power sets, ie, $V_1 = \Powset\of{V_0}$, $V_2 = \Powset\of{V_1}$, and so on. We can then define $V_{\omega}$ to be the union of all $V_n$, where $n < \omega$. We can then define $V_{\omega + 1} = \Powset\of{\omega}$, and so on. All of analysis, the field of mathematics Davoud Cheraghi once described as the ``rigorous study of infinite constructions'', is essentially about $V_{\omega + 1}$. In this way, set theory encompasses analysis, for example.

We call the hierarchy
\begin{align*}
    V_0 \subset V_1 \subset \cdots \subset V_{\omega} \subset V_{\omega + 1} \subset \cdots
\end{align*}
the \textbf{universe of sets}. Indeed, this is often represented diagrammatically as a $V$ that contains each of these $V$s, with $V_0$ at the bottom, $V_1$ above $V_0$, and so on.

There is a problem with the universe of sets: \textit{the power set operation is too strong}. That is, when we take the power set of a set, it ``throws too many elements in''. This leads to questions like the \textit{continuum hypothesis}, which asks the question of whether $\aleph_0 = \abs{\R} = \abs{V_{\omega + 1}}$ is equal to $\aleph_1$.

Something we will study quite closely in this course is a slightly different version of this universe, known as \textbf{Gödel's Constructible Universe}. This is a universe of sets that is not built up from $\emptyset$ using the power set, but using a \textit{different} operation $\Defin$, defined as follows: for a set $A$, viewing it as a first-order structure $(A; \in)$,
\begin{align*}
    \Defin(A) &:= \setst{x}{x \text{ is a 1st-order definable structure over } (A; \in) \text{ using parameters from } A}
\end{align*}
We can show that for finite sets, $\Defin$ agrees with $\Powset$. Moreover, we can show that for a countable set $A$, $\Defin(A)$ is countable as well. The point is that if we then define a universe
\begin{align*}
    \parenth{L_0 = \emptyset} \subset \parenth{L_1 = \Defin\of{L_0}} \subset \cdots \subset \parenth{L_{\omega} = \bigcup_{n < \omega} L_n} \subset \cdots
\end{align*}
we not only have agreement between $L_n$ and $V_n$ for all $n \leq \omega$, but we can build a model of ZFC in which the continuum hypothesis is true!

This is something we will explore in great detail in this course.

We will begin by reviewing the ZFC axioms.

\input{Chapters/1_Intro/1_1_ZFC.tex}
\input{Chapters/1_Intro/1_2_Another_Section.tex}